Epigraphy and onomastics in the Hesperia databank

### 2016-02-09 source/24/Epigraphy and onomastics in Hesperia databank-template.tex
fixes to Varga e Moucunill

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Epigraphy and onomastics in the Hesperia databank}
\author[barc]{Noemí Moncunill Martí \corref{first}}
\author[sorb]{Javier Velaza}
\address[sorb]{Université de Paris-Sorbonne (Paris IV)}
\address[barc]{University of Barcelona2}
\cortext[first]{Corresponding author. Email: nmoncunill@gmail.com}
\date{2015-11-15}
\begin{document}

@@

\noindent The aim of this work is to do a general overview of the computer-based epigraphic corpora, to think about their features
and advantages, taking the experience gained in the last years within the Hesperia project as a starting point.\footnote{ This paper is an output from the FFI2012-25113 project and the Senior Research Team LITTERA
(2014SGR63), on the one hand, and 655938 — ZEPHYROS — H2020-MSCA-IF-2014 project, on the other.}

The first thing that necessarily needs to be highlighted is that traditional epigraphic corpora have certain limits that
are imposed by their morphology and format. If summarizing, there are three main and most evident limitations: 
\begin{description}
\item[a] One-dimensional format. The monument's description, the textual edition, the critical apparatus and remarks are
displayed one after another, with no wider possibility to link one to another than through some indices, which by the
way are rarely detailed enough. Not even in the most extended indices, as those in CIL, the connexions between
inscriptions through such relevant data as their date, palaeography or onomastics are set clear. Obviously, those
connexions can be set by the users in their mind, but when the corpus displays thousands or tens of thousands of
inscriptions, it looks too optimistic to take for granted that most readers will have the aptitude such work takes.

@@

\end{description}
In our opinion, these limitations can be solved or, at least, diminished by a proper and efficient use of the tools IT
puts to our disposal. Although this is no longer anything of new, any attempt to use them so far has brought to an
output too close to the traditional, non-innovative epigraphic corpora, considering the features of such new tools. To
get over one-dimensional formats, an open, adaptable digital structure must be created that shall take the most out of
every single datum in the database. Besides, getting over an ephemeral usefulness requires the possibility not just to
keep adding new data, but also to connect the database with any other tool with a similar structure and to extend the
platform with any new or advisable features. Finally, the authoritative temptation must be also avoided, particularly
through systems that shall bring to users the possibility to take their own decisions in certain issues of an open
debate and, at the end of the day, to customize their own corpus.

@@

\noindent All above-mentioned principles have been taken into account in Hesperia project, its main goal being to gather all
linguistic evidences from Palaeohispanic languages, that is, pre-Roman languages in the Iberian Peninsula\footnote{
General presentations of the project can be found in \citet{orduna_implementing_????}, \citet{orduna_philology_????} and
\citet{velaza_hesperia:_2014}.}. First of all, we need to recall these are very different materials as for their quantity, quality and
reliability. Ancient Hispania left us epigraphic texts in at least four languages: Tartessian, Iberian, Celtiberian and
Lusitanian -{}-the possibility there are also some texts in Palaeobasque language is still sub iudice\footnote{
Concerning this subject, see \citet{velaza_epigrafiy_2009}.}{}-{}-, mainly written in four epichoric scripts -{}-Tartessian,
southeastern-Iberian, northeastern-Iberian and Greek-Iberian script.

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img001.jpg}
\caption{ Map of the Palaeohispanic inscriptions attested in the Iberian Peninsula.}
\label{fig:1}
\end{figure}

The current knowledge of each of these scripts is in very different stages, just as when it comes to their languages
deciphering. Along with the epigraphic texts there are also other evidences, as those from onomastics -{}-anthroponymy
and theonymy preserved in Roman inscriptions found in these territories, and toponymy from Classic sources-{}- and some
notes transmitted by several authors.

Hesperia was born as a natural successor in the digital era of the main corpus for Palaeohispanic languages, that is
Monumenta linguarum Hispanicarum, published between 1975 and 2001 by Jürgen Untermann. The project conception is due to
Javier de Hoz and its digital platform has mainly been developed by Eduardo Orduña. Nowadays it is coordinated by an
interdisciplinary and interuniversity team with researchers from the Universidad Complutense de Madrid, the Universidad
del País Vasco, the Universidad de Zaragoza and the Universidad de Barcelona. There are currently several sections open
for searching (\url{http://hesperia.ucm.es/}).

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img002.png}
\caption{Dispersal areas of Palaeohispanic inscriptions and sections available in Hesperia databank}
\label{fig:2}
\end{figure}

\begin{description}
\item[i] Partial epigraphic corpora from B zone (Iberian inscriptions from the south of France), K zone (Celtiberian
inscriptions) and L zone (Lusitanian inscriptions). All of them follow the same data form made of six tabs
corresponding to ``general features'', ``text'',
``epigraphy and paleography'', ``pictures'',
``arqueological context'' and ``bibliography''. The databases are
linked between them through a powerful engine for simple and complex searches, and pdf documents as well as maps can be
created with the resulting outcomes. Work is currently being focused on C and D zones (Iberian inscriptions in
Catalonia), which is scheduled to be open for search by the end of 2015.

@@

\item[iii] An onomastic database, with a specific search engine allowing combined searches with any of the elements typical
from the identified onomastic formulae.

@@

Beyond finishing, checking and updating the already-open sections, the team currently works on transversal fields, such
as lexicon, as well as on creating several kinds of tools, as the possibility of displaying the corpus according to
alternative readings -{}-which turns out to be essential in epigraphic contexts with a lower level of decoding, as in
Tartessian, although its applicability to other fields is still being tested-{}-.

Besides, the team has also decided to consider the potential applicability of this corpus structure to other cultures of
the western Mediterranean. With such aim, prof. Francisco Beltrán leads the AELAW project (Ancient European Languages
and Writings), with the participation of researchers from ten different countries, and that has been granted a COST
program in order to build in future a single corpus with the evidences from all fragmentary languages documented in the
western Mediterranean for the Ancient times.

@@

In this new section we have compiled all the pre-Roman personal names and divinity names attested both by direct and
indirect tradition. This represents, so far, a total of nearly 6.000 records. Names attested in direct sources are
those that can be identified in epicoric epigraphies; the second group, in its turn, contains the names that can be
identified in the so-called colonial epigraphy, as well as in literary sources. Regarding the chronology of the
epigraphic material, the oldest texts are the Iberian, which can be dated back to the 5th century BC, whereas the most
recent ones are the Latin inscriptions from the high-imperial age. Thus, the processed information can be classified
into one of these groups:

@@

As said above, the database contains about 6.000 records so far; each of them reports a divinity name and/or the whole
onomastic formula of individuals whose complete name presents at least one indigenous element. This means that each
item is devoted to a concrete person, whose onomastic formula might be composed by several indigenous names. The
information of each record is completed with a bibliographical apparatus, and the find-spot coordinates, together with
a map. Moreover, the database allows using this geographic information to create new linguistic maps, which represents
one of the main strengths offered by this new tool of the Hesperia databank.

The following could be a good example of one of the above-described records. This Latin inscription from southern Spain
mentions a woman with a Latin nomen, Aelia, followed by an Iberian cognomen, Belesiar, as well as the name of an
indigenous divinity, Betatun. Thus, a single table records an indigenous personal name together with an indigenous god.
The lower part is reserved for the bibliographical references, the geographic information and, last but not least, the
map.

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img003.png}
\caption{Personal-names and divinity-names record in Hesperia databank.}
\label{fig:3}
\end{figure}

A slightly different kind of record is conceived to compile the personal names attested in literary sources. As shown
below, in these cases the bibliographical field is merely used to mention the ancient passages in which the name is
attested; there is no geographical information, and the “observations” field is often used just to report the different
graphical variants of the name.

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img004.png}
\caption{Personal-names and divinity-names record in Hesperia databank.}
\label{fig:4}
\end{figure}

The goal of Hesperia is to provide, in the first place, an exhaustive repertory of all the Palaeohispanic names. Thus,
with the information available a map can be easily generated with all the spots where at least an indigenous name is
attested (in green); or a divinity name (in blue); or, finally, combine those maps in a single interactive map (in
green and blue), where each point is directly linked with the corresponding records.

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img005.png}
\caption{Dispersal area of indigenous personal names.}
\label{fig:5}
\end{figure}

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img006.png}
\caption{Dispersal area of indigenous divinity names}
\label{fig:}
\end{figure}

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img007.png}
\caption{Disperal areas of indigenous divinity names and personal names}
\label{fig:7}
\end{figure}

Thanks to the search engine, all these points can then be redistributed into smaller groups to draw linguistic areas or
isoglosses, on the basis, for instance, of the attestation of a well-known anthroponymic element or the dispersal area
of significant phonetic features. In the following maps it is displayed, in the first place, the attestation area of
names containing the element \emph{biur}, which clearly corresponds to the Iberian area, that is, the non indo-European half
of the Iberian peninsula; and, in the second place, it is displayed the dispersal area of the name Tancinus, which
clearly corresponds to the indo-European part.

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img008.jpg}
\caption{Dispersal area of the anthroponymic element \emph{biur}}
\label{fig:}
\end{figure}

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img009.jpg}
\caption{Dispersal area of the name Tancinuse}
\label{fig:9}
\end{figure}

@@

The following file contains information on a person’s or god’s name observed in an Iberian rock inscription in the
Pyrenees. Nevertheless, the structure of the word as well as its phonetic features show no possible interpretation in
Iberian. In consequence, other possibilities have to be considered, such as it could actually be an adaptation of a
Greek name. As a matter of fact, the same name is attested as a mythological character in literary sources (Parthenius
of Nicaea XXX 1; Ovid, Ibis 434), as well as record as a personal name in open-access databanks such as LGPN
(V5a-20577) or Trismegistos (Per 222620). Therefore, in this particular case, the linking between open-access data
could be helpful for the linguistic analysis itself.

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img010.png}
\caption{Fig. 10: record of a possible Greek name attested in an Iberian inscription.}
\label{fig:10}
\end{figure}

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Epigraphy and onomastics in the Hesperia databank}
\author[barc]{Noemí Moncunill Martí}
\author[sorb]{Javier Velaza \corref{first}}
\address[sorb]{Université de Paris-Sorbonne (Paris IV)}
\address[barc]{University of Barcelona}
\cortext[first]{Corresponding author. Email: noemi.moncunill\_marti@paris-sorbonne.fr and velaza@ub.edu}
\date{2015-11-15}
\begin{document}

@@

\noindent The aim of this work is to do a general overview of the computer-based epigraphic corpora, to think about their features
and advantages, taking the experience gained in the last years within the Hesperia project as a starting point.\footnote{ This paper is an output from the FFI2012-25113 project and the Senior Research Team LITTERA
(2014SGR63), on the one hand, and it has received funding from the European Union’s Horizon 2020 research and innovation programme under the Marie Sklodowska-Curie grant agreement No 655938, on the other.}

The first thing that necessarily needs to be highlighted is that traditional epigraphic corpora have certain limits that
are imposed by their morphology and format. If summarizing, there are three main and most evident limitations: 
\begin{description}
\item[a] One-dimensional format. The monument's description, the textual edition, the critical apparatus and remarks are
displayed one after another, with no wider possibility to link one to another than through some indices, which by the
way are rarely detailed enough. Not even in the most extended indices, as those in CIL, the connexions between
inscriptions through such relevant data as their date, palaeography or onomastics are set clear. Obviously, those
connexions can be made by the users in their mind, but when the corpus displays thousands or tens of thousands of
inscriptions, it looks too optimistic to take for granted that most readers will have the aptitude such work takes.

@@

\end{description}
In our opinion, these limitations can be solved or, at least, diminished by a proper and efficient use of the tools IT
puts to our disposal. Although this is no longer anything of new, many attempts to use them so far have brought to an
output too close to the traditional, non-innovative epigraphic corpora, considering the features of such new tools. To
get over one-dimensional formats, an open, adaptable digital structure must be created that shall take the most out of
every single datum in the database. Besides, getting over an ephemeral usefulness requires the possibility not just to
keep adding new data, but also to connect the database with any other tool with a similar structure and to extend the
platform with any new or advisable features. Finally, the authoritative temptation must be also avoided, particularly
through systems that shall bring to users the possibility to take their own decisions in certain issues of an open
debate and, at the end of the day, to customize their own corpus.

@@

\noindent All above-mentioned principles have been taken into account in Hesperia project, its main goal being to gather all
linguistic evidences from Palaeohispanic languages, that is, pre-Roman languages in the Iberian Peninsula\footnote{
General presentations of the project can be found in \citet{orduna_implementing_????}, \citet{orduna_philology_????} and
\citet{velaza_hesperia:_2014}.}. First of all, we need to recall these are very different materials as for their quantity, quality and
reliability. Ancient Hispania left us epigraphic texts in at least four languages: Tartessian, Iberian, Celtiberian and
Lusitanian --- the possibility there are also some texts in Palaeobasque language is still sub iudice\footnote{
Concerning this subject, see \citet{velaza_epigrafiy_2009}.} --- , mainly written in four epichoric writing systems --- Tartessian,
southeastern-Iberian, northeastern-Iberian and Greek-Iberian script.

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img001.jpg}
\caption{ Map of the Palaeohispanic inscriptions attested in the Iberian Peninsula.}
\label{fig:1}
\end{figure}

The current knowledge of each of these scripts is in very different stages, just as when it comes to their languages
deciphering. Along with the epigraphic texts there are also other evidences, as those from onomastics --- anthroponymy
and theonymy preserved in Roman inscriptions found in these territories, and toponymy from Classic sources --- and some
notes transmitted by several authors.

Hesperia was born as a natural successor in the digital era of the main corpus for Palaeohispanic languages, that is
\emph{Monumenta Linguarum Hispanicarum}, published between 1975 and 2001 by Jürgen Untermann. The project conception is due to
Javier de Hoz and its digital platform has mainly been developed by Eduardo Orduña. Nowadays it is coordinated by an
interdisciplinary and interuniversity team with researchers from the Universidad Complutense de Madrid, the Universidad
del País Vasco, the Universidad de Zaragoza and the Universidad de Barcelona. There are currently several sections open
for searching (\url{http://hesperia.ucm.es/}).

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img002.png}
\caption{Dispersal areas of Palaeohispanic inscriptions and sections available in Hesperia databank}
\label{fig:2}
\end{figure}

\begin{description}
\item[i] Partial epigraphic corpora from B zone (Iberian inscriptions from the south of France), K zone (Celtiberian
inscriptions) and L zone (Lusitanian inscriptions). All of them follow the same data form made of six tabs
corresponding to ``general features'', ``text'',
``epigraphy and paleography'', ``pictures'',
``archaeological context'' and ``bibliography''. The databases are
linked between them through a powerful engine for simple and complex searches, and pdf documents as well as maps can be
created with the resulting outcomes. Work is currently being focused on C and D zones (Iberian inscriptions in
Catalonia), which is scheduled to be open for search by the end of 2016.

@@

\item[iii] An onomastic database, with a specific search engine allowing combined searches with any of the elements from the identified onomastic formulae.

@@

Beyond finishing, checking and updating the already-open sections, the team currently works on transversal fields, such
as lexicon, as well as on creating several kinds of tools, as the possibility of displaying the corpus according to
alternative readings --- which turns out to be essential in epigraphic contexts with a lower level of decoding, as in
Tartessian, although its applicability to other fields is still being tested ---.

Besides, the team has also decided to consider the potential applicability of this corpus structure to other cultures of
the western Mediterranean. With such aim, prof. Francisco Beltrán leads the AELAW (Ancient European Languages
and Writings), with the participation of researchers from ten different countries to build in future a single corpus with the evidences from all fragmentary languages documented in the
western Mediterranean for the Ancient times.

@@

In this new section we have compiled all the pre-Roman personal names and divinity names attested both by direct and
indirect tradition. This represents, so far, a total of nearly 6,000 records. Names attested in direct sources are
those that can be identified in epichoric epigraphies; the second group, in its turn, contains the names that can be
identified in the so-called colonial epigraphy, as well as in literary sources. Regarding the chronology of the
epigraphic material, the oldest texts are the Iberian, which can be dated back to the 5th century BC, whereas the most
recent ones are the Latin inscriptions from the high-imperial age. Thus, the processed information can be classified
into one of these groups:

@@

As said above, the database contains about 6,000 records so far; each of them reports a divinity name and/or the whole
onomastic formula of individuals whose complete name presents at least one indigenous element. This means that each
item is devoted to a concrete person, whose onomastic formula might be composed by several indigenous names. The
information of each record is completed with a bibliographical apparatus, and the find-spot coordinates, together with
a map. Moreover, the database allows using this geographic information to create new linguistic maps, which represents
one of the main strengths offered by this new tool of the Hesperia databank.

The following (Fig.~\ref{fig:3}) could be a good example of one of the above-described records. This Latin inscription from southern Spain
mentions a woman with a Latin nomen, Aelia, followed by an Iberian cognomen, Belesiar, as well as the name of an
indigenous divinity, Betatun. Thus, a single table records an indigenous personal name together with an indigenous god.
The lower part is reserved for the bibliographical references, the geographic information and, last but not least, the
map.

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img003new.png}
\caption{Personal-names and divinity-names record in Hesperia databank.}
\label{fig:3}
\end{figure}

A slightly different kind of record is conceived to compile the personal names attested in literary sources. As shown
below (Fig.~\ref{fig:4}), in these cases the bibliographical field is merely used to mention the ancient passages in which the name is
attested; there is no geographical information, and the “observations” field is often used just to report the different
graphical variants of the name.

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img004.png}
\caption{Personal-names and divinity-names record in Hesperia databank.}
\label{fig:4}
\end{figure}

The goal of Hesperia is to provide, in the first place, an exhaustive repertory of all the Palaeohispanic names. Thus,
with the information available a map can be easily generated with all the spots where at least an indigenous name is
attested (Fig.~\ref{fig:5}); or a divinity name (Fig.~\ref{fig:6}); or, finally, combine those maps in a single interactive map (Fig.~\ref{fig:7}), where each point is directly linked with the corresponding records.

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img005.png}
\caption{Dispersal area of indigenous personal names.}
\label{fig:5}
\end{figure}

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img006.png}
\caption{Dispersal area of indigenous divinity names}
\label{fig:6}
\end{figure}

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img007.png}
\caption{Disperal areas of indigenous divinity names and personal names}
\label{fig:7}
\end{figure}

Thanks to the search engine, all these points can then be redistributed into smaller groups to draw linguistic areas or
isoglosses, on the basis, for instance, of the attestation of a well-known anthroponymic element or the dispersal area
of significant phonetic features. In the following maps it is displayed, in the first place (Fig.~\ref{fig:8}), the attestation area of
names containing the element \emph{biur}, which clearly corresponds to the Iberian area, that is, the non indo-European half
of the Iberian peninsula; and, in the second place (Fig.~\ref{fig:9}), it is displayed the dispersal area of the name Tancinus, which
clearly corresponds to the indo-European part.

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img008.jpg}
\caption{Dispersal area of the anthroponymic element \emph{biur}}
\label{fig:8}
\end{figure}

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img009.jpg}
\caption{Dispersal area of the name Tancinuse}
\label{fig:9}
\end{figure}

@@

The following file (Fig.~\ref{fig:10}) contains information on a person’s or god’s name observed in an Iberian rock inscription in the
Pyrenees. Nevertheless, the structure of the word as well as its phonetic features show no possible interpretation in
Iberian. In consequence, other possibilities have to be considered, such as it could actually be an adaptation of a
Greek name. As a matter of fact, the same name is attested as a mythological character in literary sources (Parthenius
of Nicaea XXX 1; Ovid, Ibis 434), as well as recorded as a personal name in open-access databanks such as LGPN
(V5a-20577) or Trismegistos (Per 222620). Therefore, in this particular case, the linking between open-access data
could be helpful for the linguistic analysis itself.

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img010.png}
\caption{Record of a possible Greek name attested in an Iberian inscription.}
\label{fig:10}
\end{figure}
